---
title: "Cluster annotation with Clustifyr"
author: "`r param$author`"
date: "`r format(Sys.time(), '%B, %Y')`"
geometry: "margin=2cm"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: "hide"
    highlight: "tango"
    theme: "paper"
bibliography: "`r file.path(param$path_out, 'references.bib')`"
link-citations: yes
csl: "`r file.path(param$path_to_git, 'assets/elsevier-harvard.csl')`"
css: "`r file.path(param$path_to_git, 'assets/style.css')`"
---

```{r r_options, file='../../config/configuration.R'}
#source(file.path(param$path_to_git,'config/configuration.R'))
message("Set R options")
```

```{r rmarkdown, file='../../config/rmarkdown_configuration.R'}
#source(file.path(param$path_to_git,'config/rmarkdown_configuration.R'))
message("Set Rmarkdown options")
```

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk(file.path(param$path_to_git,'scripts/export_data/text.R'))
```

```{r load_packages, warning=FALSE, message=FALSE, results='hide'}

# Load renv
renv::load(file.path(param$path_to_git,"env/basic"))

# Required libraries
library(Seurat)
library(clustifyr)
library(clustifyrdatahub)
library(patchwork)
library(magrittr)
library(ggplot2)
library(ggsci)

# Load reference datasets

if (!is.null(param$annotation_dbs_clustifyr)) {
  # Reference from clustifyrdatahub for Clustifyr
  # List https://rnabioco.github.io/clustifyrdata/articles/download_refs.html
  # Function to paste reference name and download reference dataset
  str_eval=function(x) {return(eval(parse(text=x)))}
  ref = str_eval(param$annotation_dbs_clustifyr)
  ref_name = gsub("[()]","",param$annotation_dbs_clustifyr)
  RefDataset = TRUE
} else if (!is.null(param$url_clustifyr)) {
  # Reference pulled form http://cells.ucsc.edu/ for Clustifyr
  Sys.setenv(VROOM_CONNECTION_SIZE = 131072*100)
  ref = clustifyr::get_ucsc_reference(cb_url = param$url_clustifyr, cluster_col = param$cluster_col_clustifyr)
  RefDataset = TRUE
} else {
  message("No reference dataset provided.")
  RefDataset = FALSE
}

# Remove ENS IDs 
if (all(startsWith(row.names(ref), "ENS"))) {
  row.names(ref) = gsub(".*\\|", "", row.names(ref))
}

```

# Introduction
 
```{asis, ref.label="intro_cell_annotation"}
```

Here, cluster annotation for the samples of project `r param$project_id` was performed using Clustifyr. Reference dataset for cell-type assignment with Clustifyr can be obtained from clustifyrdatahub (https://rnabioco.github.io/clustifyrdata/articles/download_refs.html) or pulled form http://cells.ucsc.edu/. The reference dataset `r ifelse(!is.null(param$annotation_dbs_clustifyr), paste0(param$annotation_dbs_clustifyr, " obtained from clustifyrdatahub was used"), "")` `r ifelse(!is.null(param$url_clustifyr), paste0("was pulled from ", param$url_clustifyr), "")`.


```{r read_rds, file='../read_data/read_rds.R', warning=FALSE}
#source(file.path(param$path_to_git,'scripts/read_data/read_rds.R'))
message("Load Seurat object from rds file")
```

```{r cell_annotation_clustifyr_fig_heights}
# Figure height for umaps
if (length(levels(sc$orig.ident))==1) {
  height_per_row = 3
} else if (length(levels(sc$orig.ident))==2) {
  height_per_row = 4
} else {
  height_per_row = 2
}
nr_rows_3cols = ceiling(length(levels(sc$orig.ident))/3)
fig_height_3plots = height_per_row * nr_rows_3cols

# Figure heights for dotplot
fig_height_dp = max(4, 4 + 0.3 * length(levels(sc$orig.ident)))

```

# Annotation of clusters
UMAPs display cells colored by Seurat clusters and cell types annotated by Clustifyr. Annotation was performed for each cluster as unit.
```{r clustifyr, eval=isTRUE(RefDataset)}
# Convert to v3; Does not work with seurat v5 at the moment
sc_v3 = scCustomize::Convert_Assay(seurat_object = sc, convert_to = "V3", assay = "RNA")

# Annotate clusters using Clustifyr with reference dataset
sc_v3 = clustify(input = sc_v3, ref_mat = ref, cluster_col = "seurat_clusters", query_genes = VariableFeatures(sc_v3))
  

# Set to annotation metadata column
sc_v3$annotation = sc_v3$type

# Set colors for cell types based on annotation 
message(paste0("Generate colors using color palette", param$col_palette_annotation))
annotation_colours_out = suppressWarnings(SetAnnotationColours(sc_v3, param$col_palette_annotation)) 
sc_v3 = annotation_colours_out[[1]]
param$col_annotation = annotation_colours_out[[2]]


# Print table with cell annotation
annotated_clusters = table(sc_v3$annotation)
knitr::kable(annotated_clusters, align="l", caption="Cell types assigned to cells") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  kableExtra::scroll_box(width="100%", height="300px")
```

<br> 
<br>
 
## Coloured by cell type 
```{r clustifyr_clusters, fig.height=fig_standard3_height, eval=isTRUE(RefDataset)}

# Visualization of Clustifyr annotation of clusters
cluster_cells = table(sc@active.ident)
cluster_labels = paste0(levels(sc@active.ident)," (", cluster_cells[levels(sc@active.ident)],")")
p1 = Seurat::DimPlot(sc_v3, reduction="umap", group.by="seurat_clusters", pt.size=param$pt_size) + 
  scale_color_manual(values=param$col_clusters, labels=cluster_labels) +
  AddStyle(title="Clusters", xlab = "UMAP 1", ylab = "UMAP 2") +
  theme(title = element_text(size = 10)) +
  NoGrid() + 
  NoLegend()
p1 = LabelClusters(p1, id="seurat_clusters")

p2 = Seurat::DimPlot(sc_v3, reduction="umap", group.by="annotation", pt.size=param$pt_size) + 
  scale_color_manual(values=param$col_annotation) +
  AddStyle(title="Cell types \n(annotation per cluster)", xlab = "UMAP 1", ylab = "UMAP 2", legend_position="right", legend_title="Cell types") +
  theme(title = element_text(size = 10)) +
  NoGrid()
p = p1 + p2
p

```

<br>
<br> 

## Coloured by cell type (per sample)
```{r clustifyr_separately, fig.height=fig_height_3plots, eval=isTRUE(RefDataset)}
# Visualization of Clustifyr annotation of cells separately per sample
p = Seurat::DimPlot(sc_v3, reduction="umap", group.by="annotation", split.by = "orig.ident", pt.size=param$pt_size, ncol = 3) + 
  scale_color_manual(values=param$col_annotation) +
  AddStyle(title="Cell types \n(annotation per cell)", xlab = "UMAP 1", ylab = "UMAP 2", legend_position="right", legend_title="Cell types") +
  theme(title = element_text(size = 10)) +
  NoGrid()
p
```

<br> 

## Fraction of sample per cell type
```{r samples_per_celltype, fig.height=6}
# Set 
cell_samples = levels(sc_v3$orig.ident)
cell_celltype = unique(sc_v3$annotation)

# Make count table
tbl = dplyr::count(sc_v3[[c("orig.ident", "annotation")]], orig.ident, annotation) %>% tidyr::pivot_wider(names_from="annotation", values_from=n, values_fill=0) %>% as.data.frame()
rownames(tbl) = paste0(tbl[,"orig.ident"])
tbl[,"orig.ident"] = NULL

tbl_bar = tbl[cell_samples, , drop=FALSE] %>% 
  tibble::rownames_to_column(var="Sample") %>%
  tidyr::pivot_longer(cols = tidyselect::all_of(cell_celltype), names_to="Celltype", values_to="Counts")
tbl_bar$Counts = as.numeric(tbl_bar$Counts)
tbl_bar$Celltype <- factor(tbl_bar$Celltype,levels = cell_celltype)
tbl_bar$Sample <- factor(tbl_bar$Sample, levels = cell_samples)

# Plot counts
p1 = ggplot(tbl_bar, aes(x=Celltype, y=Counts, fill=Sample)) + 
  geom_bar(stat="identity" ) +
  AddStyle(title="",
           fill=param$col_samples,
           legend_title="Sample",
           legend_position="bottom") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Make table with percentages
tbl_perc = round(t(tbl) / colSums(tbl) * 100, 2) %>% t() %>% as.data.frame()

tbl_bar_perc = tbl_perc[cell_samples, , drop=FALSE] %>% 
  tibble::rownames_to_column(var="Sample") %>%
  tidyr::pivot_longer(cols = tidyselect::all_of(cell_celltype), names_to="Celltype", values_to="Percentage")
tbl_bar_perc$Percentage = as.numeric(tbl_bar_perc$Percentage)
tbl_bar_perc$Celltype <- factor(tbl_bar_perc$Celltype,levels = cell_celltype)
tbl_bar_perc$Sample <- factor(tbl_bar_perc$Sample, levels = cell_samples)

# Plot percentages
p2 = ggplot(tbl_bar_perc, aes(x=Celltype, y=Percentage, fill=Sample)) + 
  geom_bar(stat="identity" ) +
  AddStyle(title="",
           fill=param$col_samples,
           legend_title="",
           legend_position="") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p = p1 + p2 + patchwork::plot_layout(guides = "collect") & theme(legend.position = 'bottom')
p
```

<br>
<br>

## Fraction of cell types per sample
```{r celltype_per_sample, fig.height=6}
# Plot counts
p1 = ggplot(tbl_bar, aes(x=Sample, y=Counts, fill=Celltype)) + 
  geom_bar(stat="identity" ) +
  AddStyle(title="", xlab = "",
           fill=param$col_annotation,
           legend_title="",
           legend_position="") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# Calculate percentages
tbl_perc = round(t(tbl) / colSums(t(tbl)) * 100, 2) %>% t() %>% as.data.frame()
tbl_perc = round(tbl / rowSums(tbl) * 100, 2) %>% as.data.frame()

# Plot percentages
tbl_bar_perc = tbl_perc[cell_samples, , drop=FALSE] %>% 
  tibble::rownames_to_column(var="Sample") %>%
  tidyr::pivot_longer(cols = tidyselect::all_of(cell_celltype), names_to="Cell_type", values_to="Percentage")
tbl_bar_perc$Percentage = as.numeric(tbl_bar_perc$Percentage)
tbl_bar_perc$Cell_type <- factor(tbl_bar_perc$Cell_type,levels = cell_celltype)
tbl_bar_perc$Sample <- factor(tbl_bar_perc$Sample, levels = cell_samples)

p2 = ggplot(tbl_bar_perc, aes(x=Sample, y=Percentage, fill=Cell_type)) + 
  geom_bar(stat="identity" ) +
  AddStyle(title="", xlab = "",
           legend_position="right",
           fill=param$col_annotation) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p = p1 + p2 + patchwork::plot_layout(guides = "collect") 
p
```

<br>

```{r dotplot_celltypes_per_sample, fig.height=fig_height_dp}
p = ggplot(data = tbl_bar_perc, aes(x = Cell_type, y = Sample)) +
  geom_point(aes(color = Percentage, size = ifelse(Percentage==0, NA, Percentage))) +
  labs(col="Percentage", size="") + 
  scale_colour_gradient2(low="navy", mid="steelblue", high="darkgoldenrod1") +
  #viridis::scale_color_viridis(option = "D") +
  AddStyle(title="", ylab="", legend_position="bottom") + 
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=.5)) + 
  guides(size=guide_legend(order=1))

p
```

<br>
<br>

```{r convert object, eval=isTRUE(RefDataset)}
 sc = scCustomize::Convert_Assay(seurat_object = sc_v3, convert_to = "V5", assay = "RNA")
```

```{r, child='../export_data/standard_export.Rmd'}
```

<br>

```{r, child='../export_data/appendix.Rmd'}
```



