
<br>

# Parameters and software versions
The following parameters were used to run the workflow.  

<details>
  <summary>Parameters</summary>
  
```{r parameters_table}
out = ScrnaseqParamsInfo(params=param)

knitr::kable(out, align="l") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover"), full_width=FALSE, position="left")
```
</details>

This report was generated using the `r param$scriptname` script. Software versions were collected at run time. 

<details>
  <summary>Software versions</summary>
  
```{r versions, message=FALSE}
out = ScrnaseqSessionInfo(param$path_to_git)

knitr::kable(out, align="l") %>% 
  kableExtra::kable_styling(bootstrap_options=c("striped", "hover"))
```

</details>

<br>

# Credits and References
This Workflow was developed as module of the sc_analysis workflow at the [Research Core Unit Genomics, Hannover Medical School](https://www.mhh.de/en/genomics) (Hannover, Germany). Seurat Vignettes were initially used as templates (`r knitcitations::citet(citation("Seurat")[1:2])`). 

```{r bib, message=FALSE}
# Writes knitcitations references to references.bib file.
knitcitations::write.bibtex(file=file.path(param$path_out, "references.bib"))
```